Astrophysical calibration of gravitational-wave detectors

### 2016-03-03 docs/cosmic_cal.tex
Some minor changes based on proofs received

\begin{abstract}
  We investigate a method to assess the validity of gravitational-wave detector calibration 
through the use of gamma-ray bursts as standard
  sirens. Such signals, as measured via gravitational-wave
  observations, provide an estimated luminosity distance that is
  subject to uncertainties in the calibration of the data.  If a host
  galaxy is identified for a given source then its redshift can be
  combined with current knowledge of the cosmological
  parameters yielding the true luminosity distance.  This will then allow
  a direct comparison with the estimated value and can validate the accuracy of the original 
calibration. We use simulations of individual detectable gravitational-wave signals from binary 
neutron star (BNS) or neutron star-black hole (NSBH) systems, which we assume to be found in 
coincidence with short gamma-ray bursts, to estimate any discrepancy in the overall scaling of the 
calibration for detectors in the Advanced LIGO and Advanced Virgo network. We find that the 
amplitude scaling of the calibration for the LIGO instruments could on average be confirmed to 
within $\sim 10\%$ for a BNS source within 100\,Mpc. This result is largely independent of the 
current detector calibration method and gives an uncertainty that is competitive with that expected 
in the current calibration procedure. Confirmation of the calibration accuracy to within 
$\sim 20\%$ can be found with BNS sources out to $\sim 500$\,Mpc.
\end{abstract}

@@

% acronym definitions
\acrodef{GW}[GW]{gravitational-wave}
\acrodef{NS}[NS]{neutron star}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{MWEG}[MWEG]{Milky Way equivalent galaxies}
\acrodef{BBH}[BBH]{binary black hole}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{NSBH}[NSBH]{neutron star-black hole}
\acrodef{SNR}[SNR]{signal-to-noise ratio}
\acrodef{GRB}[GRB]{gamma-ray burst}
\acrodef{sGRB}[sGRB]{short gamma-ray burst}
\acrodef{CBC}[CBC]{compact binary coalescence}
\acrodef{EM}[EM]{electro-magnetic}
\acrodef{aLIGO}[aLIGO]{Advanced LIGO}
\acrodef{AdV}[AdV]{Advanced Virgo}
\acrodef{PSD}[PSD]{power spectral density}
\acrodef{PDF}[PDF]{probability distribution function}
\acrodef{MCMC}[MCMC]{Markov chain Monte Carlo}
\acrodef{ET}[ET]{Einstein Telescope}
\acrodef{XRT}[XRT]{x-ray telescope}

@@

% introduce GWs and BNS systems
It is expected that the advanced generation of interferometric \ac{GW}
detectors will detect waves emitted from $O(10s)$ of \acp{CBC} per year
\cite{2010CQGra..27q3001A}. One such class of these cataclysmic
events, the inspiral and merger of \ac{BNS} systems will be detected out to a
maximum range (horizon distance) of $\approx 450$ Mpc. Assuming the
current best estimates for the cosmological parameters, this is equivalent to a
redshift $z\approx 0.1$. As noted by~\citet{1986Natur.323..310S}, \ac{CBC}
systems can be used as cosmological distance markers, otherwise known as
``standard sirens'' (analogous to \ac{EM} standard candles). The nature of
these sources will allow us to estimate, amongst other parameters, their
luminosity distance without the need for calibration against the cosmic
distance ladder.

% BNS systems are GRBs
It is considered likely that the merger of \ac{BNS} and/or \ac{NSBH} systems,
in addition to emitting detectable \acp{GW}, is also the mechanism for
producing \acp{sGRB} \cite{1992ApJ...395L..83N}. In this scenario these events
produce tightly beamed \ac{EM} emission parallel to the orbital angular
momentum vector of the system. Additional evidence for the coincidence of
\acp{sGRB} with \ac{GW} events are the estimated astrophysical rates of both
phenomena. Observations of \acp{sGRB} give an inferred rate of \ac{sGRB}
producing \ac{CBC} mergers of $8\times 10^{-9}$--$1.1\times
10^{-6}$\,Mpc$^{-3}$\,yr$^{-1}$~\cite{2012MNRAS.425.2668C} after accounting for
beaming effects. This can be compared to independently obtained estimates of
\ac{BNS} merger rates of $10^{-8}$--$10^{-5}$\,Mpc$^{-3}$\,yr$^{-1}$
\cite{2010CQGra..27q3001A}, or \ac{NSBH} merger rates of
$6\,\times\,10^{-10}$--$1\,\times\,10^{-6}$\,Mpc$^{-3}$\,yr$^{-1}$
\cite{2010CQGra..27q3001A}\footnote{However, the recent work of
\cite{2015MNRAS.448..928K}, which studied known galactic \ac{BNS} systems in
more detail suggest a rate approximately five times lower.}.

If a single \ac{GW} event is observed in coincidence with a \ac{sGRB} it may be
possible to identify the host galaxy of the source.  With this information a
spectroscopic redshift can be very accurately obtained (see values quoted in
e.g.\ \cite{0004-637X-664-2-1000} and references therein). With knowledge of
the redshift, using current best estimates for the Hubble constant and other
cosmological parameters, the true luminosity distance can be estimated to $\sim
1\%$ accuracy~\cite{2015arXiv150201589P}.  The \ac{GW} measurement
acts as a standard siren also giving us a direct measurement of the luminosity
distance to the source. The accuracy of such a measurement depends on a number
of factors including the accuracy with which the \ac{GW} detector has been
calibrated.  Hence, by comparison with the distance estimate from the \ac{sGRB}
we can re-calibrate (or validate) the existing experimentally obtained
calibration.

% What do we do in this paper
In this paper we investigate the feasibility of this approach for single coincident 
\ac{GW}--\ac{sGRB} events and establish the validation power of such a calibration technique.  In 
Sec.~\ref{sec:calibration} we briefly summarize the existing experimental technique for \ac{GW} 
detector calibration and its expected accuracy.  We then review the concept of \ac{GW} standard 
sirens and their proposed coincident \ac{EM} signatures, the \acp{sGRB} in
Secs.~\ref{sec:sirens} and~\ref{sec:GRB}. In Secs.~\ref{sec:analysis} and \ref{sec:results} we 
discuss our analysis method and the results, and in Sec.~\ref{sec:discussion} we 
conclude.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Gravitational wave detector calibration\label{sec:calibration}}

@@

% Describe our model for the calibration
The \ac{GW} strain is measured through the differential arm length, $\Delta L$,
changes of the interferometer via $h(f,t) = \Delta L(f,t) / L$, where $L$ is
the full arm length. Calibration is required to relate the actual measured
interferometer error signal output $e(f, t)$ to $\Delta L$.  This relation is
known as the length response function, $R(f)$, defined such that
%
\begin{equation}
\Delta L(f,t) = R(f) e(f, t),
\end{equation}
%
where we assume $R$ varies only slowly in time (in comparison to transient
signal time-scales). Calculation of $R$ requires the measurement of various
functions within a control feedback loop (see \cite{2010NIMPA.624..223A}),
each of which are subject to measurement uncertainties. In this study
we will assume an estimate of $R$ is available (although in theory we could
take on the role of estimating $R$ itself), but that it differs from the truth
through some unknown scale factor, $\scf$, so that at a particular time we have
%
\begin{equation}\label{eq:scalefactor}
\scf h(f) = h_{\rm m}(f) = \frac{R(f) e(f)}{L},
\end{equation}
%
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain
(i.e.\ the measured length response function $R(f)$ is related to the
true length response function $R_{\rm true}(f)$ via $R(f) = \scf R_{\rm
true}(f)$). With this definition it means that if $\scf > 1$ then a signal
would appear to have a larger amplitude (i.e.\ for a given system,
would seem closer) than in reality\footnote{However, the \ac{SNR} would be
the correct value as the signal and the noise will both contain the same scale
factor.}, whereas if $\scf < 1$ it would appear to have a lower amplitude than
reality.  In this analysis we will simplify the situation by assuming that
$\scf$ is a constant with respect to frequency (and is real, so has no phase
component), but in future studies that assumption could be dropped and $\scf$
could take some functional form, or piece-wise fit, with respect to $f$ (e.g.\
in a similar way to the method of \cite{2013PhRvD..88h4044L} used for fitting
differences in \ac{PSD} estimates).

@@

% Standard siren basics
The idea of \ac{GW} standard sirens is directly analogous to the concept of
standard candles in \ac{EM} astronomy and was first proposed
in~\cite{1986Natur.323..310S}. Unlike the primary \ac{EM} standard candle
event, Type Ia Supernovae, the measured luminosity of a \ac{CBC} in \acp{GW} is
not only a function of distance. It also depends upon the chirp mass (a
function of the component masses) and the binary orientation with respect to
the global network of \ac{GW} detectors. However, measurement of the
phase evolution of such an event allows accurate determination of the chirp
mass\footnote{In reality it is the redshifted chirp mass that is measured.}.
Timing information from the different signal arrival times at each detector in
the network allows some level of sky position
determination~\cite{2014ApJ...795..105S}.  Finally, amplitude variation between
differently oriented interferometers allows some (weaker) level of
determination of the binary system orientation.

% The classic use of sirens
Standard sirens are clearly a very powerful tool for \ac{GW} cosmology since
they give us a direct measure of the absolute luminosity distance to sources.
This is distinct from Type Ia Supernovae standard candles that only provide
relative luminosity distance measures and require calibration via other methods
as part of the cosmological distance ladder. The use of \ac{GW} events with
\ac{sGRB} counterparts for use as a cosmological tool was investigated
in~\cite{2010CQGra..27u5006S} and~\cite{2011PhRvD..83b3005Z} with respect to
the 3rd generation \ac{GW} interferometer, the
\ac{ET}~\cite{2010CQGra..27h4007P} and in ~\cite{2012PhRvD..85b3535T}
for the Advanced detector network. In this case, the redshift information
obtained from the \ac{sGRB} host galaxy, provides a complementary measurement
allowing each event to inform the distance-redshift relationship and hence
obtain cosmological parameter estimates. Other methods for cosmological
inference (see~\cite{1996PhRvD..53.2878F,2012PhRvD..85b3535T,
2012PhRvD..86b3502T,Messenger:2011ux,2013arXiv1312.1862M}) have been proposed
for \ac{BNS} systems that do not rely on any redshift measurements. These
instead use either the distribution of measured \acp{SNR} and assumed
form of \ac{NS} mass and spatial distributions or exploit the
features of the tidal and post-merger hyper-massive \ac{NS} stages of the
\ac{GW} waveform.

% put this in context with calibration
To put the existing work on \ac{GW} cosmology in context with the study
described in this manuscript we are essentially turning the standard
cosmological problem upside-down. We assume that the existing \ac{EM}
cosmological parameter estimation (specifically for the Hubble constant, which
is has uncertainties of $O(1\%)$~\cite{2015arXiv150201589P}) is
accurately determined and therefore a known quantity. We then assume that the
unknown in our analysis is the absolute amplitude calibration of the
interferometers in the global \ac{GW} detector network.

@@

% observing a coincident event
The most likely scenario in which a coincident \ac{GW}--\ac{sGRB} event would
be identified is through the targeted follow-up of an observed \ac{GW} or by
post-facto matching of \ac{GW} trigger lists with known (or subsequently found
\cite{2015ApJS..217....8B}) \ac{sGRB} events.  The likelihood of being able to
follow-up \ac{GW} events using gamma-ray telescopes with low enough latency to
catch a \ac{sGRB} is low.  Compounding this issue is the relatively large
\ac{GW} sky error-box giving a field-of-view for the \ac{EM} observatories to
search spanning $\sim 100$'s of square degrees (e.g.\
\cite{2011CQGra..28j5021F,2013arXiv1304.0670L,2014ApJ...795..105S}).
For the scenario of a \ac{GW}-followup of an observed \ac{sGRB}, the
merger time for \ac{BNS}/\ac{NSBH} systems will be estimated from the \ac{sGRB}
to within a few seconds (see e.g.\ 340 the discussion in Sec.\ 2.2 of
\cite{2012ApJ...760...12A}). This makes the follow-up search less
computationally expensive since it is performed over a
smaller range of data using potentially fewer numbers of waveform templates.
This computational saving enables the use of a more computationally expensive
multi-detector coherent scheme rather than the cheaper coincidence methods used
in the untargeted searches.

% discuss the inclination angle issue
A fortunate consequence of a joint \ac{GW}--\ac{sGRB} observation will be the
fact that in order for such an event to be observed, the \ac{BNS}/\ac{NSBH}
system must have had its orbital angular momentum vector pointing towards (or
away) from the detector.  The actual inclination angle of the system, defined
as the angle between the orbital angular momentum vector and the line of sight,
must be $<$ half of the \ac{sGRB} beaming angle.  This prerequisite property
limits us to systems that are approximately ``face-on'' and therefore
effectively optimally oriented. At given distances this biases us to  higher
\ac{SNR} signals but it is more meaningful to think of this as an increase in
the sensitivity range of the detectors. However, as discussed earlier, the
property of beaming severely impacts the probable rate of such joint
observations.

>>>>

\begin{abstract}
  We investigate a method to assess the validity of gravitational-wave detector calibration 
through the use of gamma-ray bursts as standard
  sirens. Such signals, as measured via gravitational-wave
  observations, provide an estimated luminosity distance that is
  subject to uncertainties in the calibration of the data.  If a host
  galaxy is identified for a given source then its redshift can be
  combined with current knowledge of the cosmological
  parameters yielding the true luminosity distance.  This will then allow
  a direct comparison with the estimated value and can validate the accuracy of the original 
calibration. We use simulations of individual detectable gravitational-wave signals from binary 
neutron star (BNS) or neutron star-black hole systems, which we assume to be found in 
coincidence with short gamma-ray bursts, to estimate any discrepancy in the overall scaling of the 
calibration for detectors in the Advanced LIGO and Advanced Virgo network. We find that the 
amplitude scaling of the calibration for the LIGO instruments could on average be confirmed to 
within $\sim 10\%$ for a BNS source within 100\,Mpc. This result is largely independent of the 
current detector calibration method and gives an uncertainty that is competitive with that expected 
in the current calibration procedure. Confirmation of the calibration accuracy to within 
$\sim 20\%$ can be found with BNS sources out to $\sim 500$\,Mpc.
\end{abstract}

@@

% acronym definitions
\acrodef{GW}[GW]{gravitational-wave}
\acrodef{NS}[NS]{neutron star}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{MWEG}[MWEG]{Milky Way equivalent galaxies}
\acrodef{BBH}[BBH]{binary black hole}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{NSBH}[NSBH]{neutron star-black hole}
\acrodef{SNR}[SNR]{signal-to-noise ratio}
\acrodef{GRB}[GRB]{gamma-ray burst}
\acrodef{sGRB}[sGRB]{short gamma-ray burst}
\acrodef{CBC}[CBC]{compact binary coalescence}
\acrodef{EM}[EM]{electromagnetic}
\acrodef{aLIGO}[aLIGO]{Advanced LIGO}
\acrodef{AdV}[AdV]{Advanced Virgo}
\acrodef{PSD}[PSD]{power spectral density}
\acrodef{PDF}[PDF]{probability distribution function}
\acrodef{MCMC}[MCMC]{Markov chain Monte Carlo}
\acrodef{ET}[ET]{Einstein Telescope}
\acrodef{XRT}[XRT]{x-ray telescope}

@@

% introduce GWs and BNS systems
It is expected that the advanced generation of interferometric \ac{GW}
detectors will detect waves emitted from $O(10s)$ of \acp{CBC} per year
\cite{2010CQGra..27q3001A}. One such class of these cataclysmic
events, the inspiral and merger of \ac{BNS} systems will be detected out to a
maximum range (horizon distance) of $\approx 450$ Mpc. Assuming the
current best estimates for the cosmological parameters, this is equivalent to a
redshift $z\approx 0.1$. As noted by~\citet{1986Natur.323..310S}, \ac{CBC}
systems can be used as cosmological distance markers, otherwise known as
``standard sirens'' (analogous to \ac{EM} standard candles). The nature of
these sources will allow us to estimate, among other parameters, their
luminosity distance without the need for calibration against the cosmic
distance ladder.

% BNS systems are GRBs
It is considered likely that the merger of \ac{BNS} and/or \ac{NSBH} systems,
in addition to emitting detectable \acp{GW}, is also the mechanism for
producing \acp{sGRB} \cite{1992ApJ...395L..83N}. In this scenario these events
produce tightly beamed \ac{EM} emission parallel to the orbital angular
momentum vector of the system. Additional evidence for the coincidence of
\acp{sGRB} with \ac{GW} events is the estimated astrophysical rates of both
phenomena. Observations of \acp{sGRB} give an inferred rate of \ac{sGRB}
producing \ac{CBC} mergers of $8\times 10^{-9}$--$1.1\times
10^{-6}$\,Mpc$^{-3}$\,yr$^{-1}$~\cite{2012MNRAS.425.2668C} after accounting for
beaming effects. This can be compared to independently obtained estimates of
\ac{BNS} merger rates of $10^{-8}$--$10^{-5}$\,Mpc$^{-3}$\,yr$^{-1}$
\cite{2010CQGra..27q3001A}, or \ac{NSBH} merger rates of
$6\,\times\,10^{-10}$--$1\,\times\,10^{-6}$\,Mpc$^{-3}$\,yr$^{-1}$
\cite{2010CQGra..27q3001A}\footnote{However, the recent work of
\cite{2015MNRAS.448..928K}, which studied known galactic \ac{BNS} systems in
more detail, suggests a rate approximately five times lower.}.

If a single \ac{GW} event is observed in coincidence with a \ac{sGRB} it may be
possible to identify the host galaxy of the source.  With this information a
spectroscopic redshift can be very accurately obtained (see values quoted in
e.g.\ \cite{0004-637X-664-2-1000} and references therein). With knowledge of
the redshift, using current best estimates for the Hubble constant and other
cosmological parameters, the true luminosity distance can be estimated to $\sim
1\%$ accuracy~\cite{2015arXiv150201589P}.  The \ac{GW} measurement
acts as a standard siren also giving us a direct measurement of the luminosity
distance to the source. The accuracy of such a measurement depends on a number
of factors including the accuracy with which the \ac{GW} detector has been
calibrated.  Hence, by comparison with the distance estimate from the \ac{sGRB}
we can recalibrate (or validate) the existing experimentally obtained
calibration.

% What do we do in this paper
In this paper we investigate the feasibility of this approach for single coincident 
\ac{GW}--\ac{sGRB} events and establish the validation power of such a calibration technique.  In 
Sec.~\ref{sec:calibration} we briefly summarize the existing experimental technique for \ac{GW} 
detector calibration and its expected accuracy.  We then review the concept of \ac{GW} standard 
sirens and their proposed coincident \ac{EM} signatures, the \acp{sGRB} in
Secs.~\ref{sec:sirens}--\ref{sec:GRB}. In Secs.~\ref{sec:analysis}--\ref{sec:results} we 
discuss our analysis method and the results, and in Sec.~\ref{sec:discussion} we 
conclude.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Gravitational-wave detector calibration\label{sec:calibration}}

@@

% Describe our model for the calibration
The \ac{GW} strain is measured through the differential arm length, $\Delta L$,
changes of the interferometer via $h(f,t) = \Delta L(f,t) / L$, where $L$ is
the full arm length. Calibration is required to relate the actual measured
interferometer error signal output $e(f, t)$ to $\Delta L$.  This relation is
known as the length response function, $R(f)$, defined such that
%
\begin{equation}
\Delta L(f,t) = R(f) e(f, t),
\end{equation}
%
where we assume $R$ varies only slowly in time (in comparison to transient
signal time scales). Calculation of $R$ requires the measurement of various
functions within a control feedback loop (see \cite{2010NIMPA.624..223A}),
each of which are subject to measurement uncertainties. In this study
we assume an estimate of $R$ is available (although in theory we could
take on the role of estimating $R$ itself), but that it differs from the truth
through some unknown scale factor, $\scf$, so that at a particular time we have
%
\begin{equation}\label{eq:scalefactor}
\scf h(f) = h_{\rm m}(f) = \frac{R(f) e(f)}{L},
\end{equation}
%
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain
(i.e.\ the measured length response function $R(f)$ is related to the
true length response function $R_{\rm true}(f)$ via $R(f) = \scf R_{\rm
true}(f)$). With this definition it means that if $\scf > 1$ then a signal
would appear to have a larger amplitude (i.e.\ for a given system,
would seem closer) than in reality\footnote{However, the \ac{SNR} would be
the correct value as the signal and the noise will both contain the same scale
factor.}, whereas if $\scf < 1$ it would appear to have a lower amplitude than
reality.  In this analysis we simplify the situation by assuming that
$\scf$ is a constant with respect to frequency (and is real, so has no phase
component), but in future studies that assumption could be dropped and $\scf$
could take some functional form, or piecewise fit, with respect to $f$ [e.g.\
in a similar way to the method of \cite{2013PhRvD..88h4044L} used for fitting
differences in \ac{PSD} estimates].

@@

% Standard siren basics
The idea of \ac{GW} standard sirens is directly analogous to the concept of
standard candles in \ac{EM} astronomy and was first proposed
in~\cite{1986Natur.323..310S}. Unlike the primary \ac{EM} standard candle
event, type Ia supernovae, the measured luminosity of a \ac{CBC} in \acp{GW} is
not only a function of distance. It also depends upon the chirp mass (a
function of the component masses) and the binary orientation with respect to
the global network of \ac{GW} detectors. However, measurement of the
phase evolution of such an event allows accurate determination of the chirp
mass\footnote{In reality it is the redshifted chirp mass that is measured.}.
Timing information from the different signal arrival times at each detector in
the network allows some level of sky position
determination~\cite{2014ApJ...795..105S}.  Finally, amplitude variation between
differently oriented interferometers allows some (weaker) level of
determination of the binary system orientation.

% The classic use of sirens
Standard sirens are clearly a very powerful tool for \ac{GW} cosmology since
they give us a direct measure of the absolute luminosity distance to sources.
This is distinct from type Ia supernovae standard candles that only provide
relative luminosity distance measures and require calibration via other methods
as part of the cosmological distance ladder. The use of \ac{GW} events with
\ac{sGRB} counterparts for use as a cosmological tool was investigated
in~\cite{2010CQGra..27u5006S}--\cite{2011PhRvD..83b3005Z} with respect to
the third generation \ac{GW} interferometer, the
Einstein Telescope~\cite{2010CQGra..27h4007P} and in ~\cite{2012PhRvD..85b3535T}
for the Advanced detector network. In this case, the redshift information
obtained from the \ac{sGRB} host galaxy provides a complementary measurement
allowing each event to inform the distance-redshift relationship and hence
obtain cosmological parameter estimates. Other methods for cosmological
inference (see~\cite{1996PhRvD..53.2878F,2012PhRvD..85b3535T,
2012PhRvD..86b3502T,Messenger:2011ux,2013arXiv1312.1862M}) have been proposed
for \ac{BNS} systems that do not rely on any redshift measurements. These
instead use either the distribution of measured \acp{SNR} and assumed
form of \ac{NS} mass and spatial distributions or exploit the
features of the tidal and postmerger hypermassive \ac{NS} stages of the
\ac{GW} waveform.

% put this in context with calibration
To put the existing work on \ac{GW} cosmology in context with the study
described in this manuscript we are essentially turning the standard
cosmological problem upside down. We assume that the existing \ac{EM}
cosmological parameter estimation [specifically for the Hubble constant, which
has uncertainties of $O(1\%)$~\cite{2015arXiv150201589P}] is
accurately determined and therefore a known quantity. We then assume that the
unknown in our analysis is the absolute amplitude calibration of the
interferometers in the global \ac{GW} detector network.

@@

% observing a coincident event
The most likely scenario in which a coincident \ac{GW}--\ac{sGRB} event would
be identified is through the targeted follow-up of an observed \ac{GW} or by
post-facto matching of \ac{GW} trigger lists with known (or subsequently found
\cite{2015ApJS..217....8B}) \ac{sGRB} events.  The likelihood of being able to
follow-up \ac{GW} events using gamma-ray telescopes with low enough latency to
catch a \ac{sGRB} is low.  Compounding this issue is the relatively large
\ac{GW} sky error-box giving a field-of-view for the \ac{EM} observatories to
search spanning $\sim 100$'s of square degrees (e.g.\
\cite{2011CQGra..28j5021F,2013arXiv1304.0670L,2014ApJ...795..105S}).
For the scenario of a \ac{GW}-followup of an observed \ac{sGRB}, the
merger time for \ac{BNS}/\ac{NSBH} systems will be estimated from the \ac{sGRB}
to within a few seconds (see e.g.\ the discussion in Sec.\ 2.2 of
\cite{2012ApJ...760...12A}). This makes the follow-up search less
computationally expensive since it is performed over a
smaller range of data using potentially fewer numbers of waveform templates.
This computational saving enables the use of a more computationally expensive
multi-detector coherent scheme rather than the cheaper coincidence methods used
in the untargeted searches.

% discuss the inclination angle issue
A fortunate consequence of a joint \ac{GW}--\ac{sGRB} observation will be the
fact that in order for such an event to be observed, the \ac{BNS}/\ac{NSBH}
system must have had its orbital angular momentum vector pointing towards (or
away) from the detector.  The actual inclination angle of the system, defined
as the angle between the orbital angular momentum vector and the line of sight,
must be $<$ half of the \ac{sGRB} beaming angle.  This prerequisite property
limits us to systems that are approximately ``face-on'' and therefore
effectively optimally oriented. At given distances this makes us biased towards higher
\ac{SNR} signals but it is more meaningful to think of this as an increase in
the sensitivity range of the detectors. However, as discussed earlier, the
property of beaming severely impacts the probable rate of such joint
observations.
